---
title: "The Exponential Distribution & Central Limit Theorem"
author: "Renaud Dufour"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(ggplot2)
```

# Overview

In this report, we investigate the exponential distribution and compare it with
the Central Limit Theorem (CLT). As a reminder, the CLT states that the distribution of averages of iid variables
(properly normalized) becomes that of a standard normal as the sample size $n$ increases,
in other words, $\bar{X_n}$ is approximately $N(\mu,\sigma^2/n)$.

# Simulation

We want to investigate how the arithmetic mean of exponentially distributed
iid variables converges to a normal distribution.
The exponential distribution is the probability distribution that describes the
time between events in a Poisson process $P(X=x, \lambda) = \lambda \exp{-\lambda x}$.
$\lambda$ is the rate parameter. The mean of the exponential distribution is
$\mu = 1/\lambda$ and the standard deviation is also $\sigma = 1/\lambda$.
This distribution can be simulated in R with ```rexp(n, lambda)```.
For all of the simulations, **we set $\lambda = 0.2$**, and we investigate the
distribution of **averages of 40 exponentials**.
We will consider **a thousand simulations**.  

The following code generates the data. An histogram of the simulated data, along with
the theoretical distribution, are shown in figure A1 of the Appendix.

```{r}
set.seed(2534)  # for reproducibility
nosim  <- 1000  # number of simulated averages
lambda <- 0.2   # rate parameter
data  <- data.frame( x = rexp(nosim * 40, lambda))
```

Then, we compute the distribution of averages $\bar{X}_{n=40}$ :

```{r}
averages <- data.frame(x = apply(matrix(data$x,nosim),1,mean))
```

In the following, we will focus on the above distribution of 40 exponentially distributed iid, 
addressing the following points :

1. How does the sample mean compare to the theoretical mean of the distribution.
2. How does the sample variance compare to the theoretical variance of the distribution.
3. How does the distribution compares to a normal distribution.

# Results

## Sample mean

Considering the distribution of the mean of 40 exponentials. Because the mean is
an unbiased estimator and is consistent, the sample mean of this distribution converges
to the population mean of the initial exponential distribution.
Thereby the theoretical mean of this distribution is $\mu = 1/\lambda = 5$.  

The simulated sample mean is :
```{r, include = TRUE}
sample.mean <- mean(averages$x)
sample.mean
```
which is indeed close to $\mu = 5$.

In order to illustrate this, the distribution is plotted below as a density plot. 
The position of the sample mean is indicated by a vertical red line.

```{r, fig.width=3, fig.height=3}
ggplot(averages, aes(x = x)) + 
geom_density() +
geom_vline(xintercept=sample.mean, size = 1, color = 'red') 
```

**Figure 1: distribution of averages of 40 exponential random variables with rate 
$\lambda = 0.2$. The vertical red line indicates the sample mean whose theoretical
value is the mean of the original exponential distribution $\mu = 1/\lambda = 5$.**

## Sample variance

The theoretical value for the variance of the distribution of averages is given by the variance
of the original population $\sigma^2$ divided by the number of samples $n$ used to compute the averages :
$var(\bar{X}) = \frac{\sigma^2}{n} = \frac{1}{40 \lambda^2} = 0.625$.

The simulated sample variance is :
```{r, include = TRUE}
var(averages$x)
```
which is indeed close to the theoretical value $0.625$.

To go further in illustrating the distribution of averages, one could repeat 
the above simulation several times and plot the distribution of sample mean.
One would observe that the sample means are distributed with a mean $\mu$ and
their variance is centered around $\sigma/\sqrt{40}$.
This is shown in figure A2 of the appendix.

## Comparison to the normal distribution

In order to compare the distribution of averages to the Central Limit Theorem (CLT),
we plot below the distribution as an histogram along with a normal distribution $N(\mu,\sigma/\sqrt{n})$.

```{r, fig.width=5, fig.height=3}
mu    <- 1/lambda
sigma <- 1/lambda
ggplot(averages, aes(x = x)) + 
geom_histogram(alpha = .10, binwidth=0.1, colour = "black", aes(y = ..density..)) +
stat_function(geom = "line", fun = dnorm, arg = list(mean = mu, sd = sigma/sqrt(40)),
              size = 2, colour = "red", fill = NA)
```

**Figure 2: distribution of averages of 40 exponential random variables (rate 
$\lambda = 0.2$). The theoretical mean and standard deviation of this distribution are respectively
$\mu = 1/\lambda$ and $\sigma = 1/(\lambda\sqrt{40})$. To illustrate the central limit theorem, the distribution is compared
 to a normal distribution $N(\mu,\sigma)$.**

We can see in figure 2 that, qualitatively, the two distributions compare relatively well.
As an alternative to compare them, one can also draw a quantile-quantile plot,
 this is shown in figure A3 of the appendix. Finally, one can also perform an hypothesis 
 test, considering the null hypothesis $H_0$ : *the distribution of averages is normal with 
 mean $\mu$ and standard deviation $\sigma$*. The test statistic $Z$ and p-value are :
  
```{r}
Z <- abs((mean(averages$x)-mu)/(sigma/sqrt(40)))
p <- 2*pnorm(Z, lower.tail = FALSE)
c(Z,p)
```
We see from the very large p-value that, even considering a extreme type I error 
rate of 99%, we would still fail to reject the null hypothesis that the data is normal.
This gives a relatively strong evidence that the distribution of averages is approximatly normal.

\pagebreak

# Appendix

## Exponential distribution

```{r, fig.width=5, fig.height=3}
ggplot(data, aes(x = x)) + 
geom_histogram(alpha = .20, binwidth=0.8, colour = "black", aes(y = ..density..)) +
stat_function(fun = dexp, arg = list(rate = lambda), size = 1, colour = "red")
```

**Figure A1 : histogram of the simulated data overlaid with the theoretical 
exponential distribution (red curve, $\lambda = 0.2$)**

## Repeated simulation

```{r, fig.width=7, fig.height=2.5}
library(gridExtra)
set.seed(1332)
nosim  <- 1000
lambda <- 0.2  # rate parameter
data  <- data.frame( x = rexp(nosim * 40, lambda))

sample.mean = NULL
sample.var  = NULL
for (i in 1 : 100) {
  x = apply(matrix(rexp(nosim * 40, lambda),nosim),1,mean)
  sample.mean = c(sample.mean, mean(x))
  sample.var  = c(sample.var,  var(x))
  }

dat <- data.frame(sample.mean,sample.var)

plot1 <- ggplot(dat, aes(sample.mean)) + geom_density() +
  geom_vline(xintercept=mean(sample.mean), size = 1, color = 'red') 

plot2 <- ggplot(dat, aes(sample.var)) + geom_density() +
  geom_vline(xintercept=mean(sample.var), size = 1, color = 'red')

grid.arrange(plot1, plot2, ncol=2)
```

**Figure A2 : Distribution of sample mean (left) and sample variance (right) obtained 
by repeating the simulation 100 times. The mean sample mean and mean sample variance (vertical red lines),
are respectively ```r mean(dat$sample.mean)``` and ```r mean(dat$sample.var)```, very close to the
theoretical values $\mu = 5$ and $var(\bar{X}) = 0.625$.**

## Quantile-Quantile plot

```{r, fig.width = 3, fig.height = 2.5}
par(mar=c(4,4,0,0)+0.1,mgp=c(2,1,0))
y <- (averages$x-mu)/(sigma/sqrt(40))  # normalize the data
qqnorm(y, main = NULL)
qqline(y)
```

**Figure 4: QQ plot of the distribution of averages of 40 exponential random variables
(rescaled as $\frac{\bar{X}-\mu}{\sigma/\sqrt{40}}$). The straight line is the standard normal reference. 
From this, it is likely that the population is normally distributed.**

