---
title: "Tooth Growth Analysis"
author: "Renaud Dufour"
date: "Wednesday, January 21, 2015"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(ggplot2)
library(plyr)
library(dplyr)
```

# Overview

This report is part of the cource project of Statistical Inference on Coursera.
We analyse the R built-in **ToothGrowth** dataset.
The latter contains data about the length of odontoblasts (teeth) for guinea pigs 
for two different supplements of Vitamin C and at different dose levels.
After summarizing the dataset and performing some exploratory analysis, we will use 
hypothesis tests in order to compare tooth growth by supplement type and dose.

# Data summary and exploratory analysis

First we load the ToothGrowth dataset (renaming variables with more explicit names) :
```{r load}
data(ToothGrowth)
names(ToothGrowth) <- c("length", "supplement", "dose")
```

The dataset contains the following variables :  

* **length** : numeric, tooth length
* **supplement** : factor, supplement type (VC for orange juice or OJ for ascorbic acid).
* **dose** : numeric,	dose in milligrams (0.1, 1 or 2).  

There is in total 60 observations (10 pigs for each supplement-dose pair).
Below is a summary of the data :
```{r summary, include = TRUE}
summary(ToothGrowth)
```

And a box plot showing tooth length vs. supplement type and dose :
```{r boxplot, fig.width=6, fig.height=4}
supplement_names    <- list( 'OJ'="Orange Juice", 'VC'="Ascorbic Acid")
supplement_labeller <- function(variable,value){supplement_names[value]}

ggplot(ToothGrowth, aes(x = as.factor(dose), y = length, fill = as.factor(dose)) ) +
  facet_grid(~supplement, labeller = supplement_labeller) +
  geom_boxplot() + theme_bw() + 
  labs(title="Tooth Growth in Guinea Pigs \n",
       x = "Dose [mg]",
       y="Change in Length [mm]") + 
  guides(fill = guide_legend("Dose [mg]"))
```

**Figure 1: Tooth length as a function of supplement type and dose.**

# Comparison of tooth growth by supplement and dose (t-tests)

In the following, we analyze the impact of supplement type and dose on the tooth
growth using hypothesis testing.
For all the following tests, we assume normality of the underlying data.

First we perform a series of pairwise t.test for each supplement type.
We consider the null hypothesis $H_0$ : *"true difference in means is equal to 0"* 
versus the alternative hypothesis $H_a$ : *"true difference in means is not equal to 0"*,
Tests are two-sided with a default $\alpha = 5\%$ type 1 error rate and we consider unequal variances.
```{r pairwise t.test dose}
# Orange Juice supplement
OJ.data <- filter(ToothGrowth, supplement=="OJ")
OJ.crosstest <- with(OJ.data, pairwise.t.test(length, dose, pool.sd = FALSE, 
                                              p.adjust.method = "none"))
# Ascorbic Acid supplement
VC.data <- filter(ToothGrowth, supplement=="VC")
VC.crosstest <- with(VC.data, pairwise.t.test(length, dose, pool.sd = FALSE, 
                                              p.adjust.method = "none"))
OJ.crosstest$p.value
VC.crosstest$p.value
```

From the above results, we see that all p-values are  below $\alpha$, so we can
conclude that the supplement dose has a significant impact on tooth growth, this whatever
the type of supplement used.

We then perform another series of t-test to investigate if, for a given dose,
the orange juice is more efficient than the ascorbic Acid (as suggested by looking at figure 1).
We consider in this case the null hypothesis $H_0$ : *"true difference in means is equal to 0"* 
versus the alternative hypothesis $H_a$ : *"true difference in means is greater than 0"*,
Tests are one sided with a default $\alpha = 5\%$ type 1 error rate and we consider unequal variances.

```{r pairwise t.test supplement}

getp <- function(x) {
  a <- x[x$supplement == "OJ","length"]
  b <- x[x$supplement == "VC","length"]
  t.test(a, b, alternative = "greater", var.equal = FALSE)$p.value
  }

suptest <- ddply(ToothGrowth, .(dose), getp)
names(suptest) <- c("dose", "p-value")
round(suptest,4)
```

The results above show, for each dose, the obtained p-value.
We observe that, as far as low and medium doses are concerned, there is a significant increase
in mean tooth length when using Orange Juice compared to Ascorbic acid. For the higher dose however,
 the difference between the two supplement types is not significant.

# Conclusion

In this report we have analyzed the **ToothGrowth** dataset containing data about the length of odontoblasts (teeth) for guinea pigs for two different supplements of Vitamin C and at different dose levels.
Assuming normality of the underlying data, We have performed a series of t-test showing that :

* For a given supplement type, there is a significant increase in tooth length for doses of 0.5, 1 and 2 mg
(two-sided tests, $\alpha = 5\%$, unequal variances).
* For a given dose, the Orange juice appears globally more efficient compared to the ascorbic acid
(significantly larger tooth length for doses of 0.5 and 1 mg) (two-sided t-tests, $\alpha = 5\%$, unequal variances).

It is to be noted that the analysis was performed based only on the techniques from class (up to week 3 lectures).
Therefore we made no correction of the p-values when doing multiple comparisons.
The changes in the results obtained by considering equal variances or corrected p-values are reported in the appendix.

\pagebreak

# Appendix

## t-tests with equal variances

Below are the results of pairwise t.test for each supplement type, considering equal variances.
The null hypothesis $H_0$ : *"true difference in means is equal to 0"* 
versus $H_a$ : *"true difference in means is not equal to 0"*,
Tests are two-sided with a default $\alpha = 5\%$ type 1 error rate.
```{r}
# Orange Juice supplement
OJ.data <- filter(ToothGrowth, supplement=="OJ")
OJ.crosstest <- with(OJ.data, pairwise.t.test(length, dose, pool.sd = TRUE, 
                                              p.adjust.method = "none"))
# Ascorbic Acid supplement
VC.data <- filter(ToothGrowth, supplement=="VC")
VC.crosstest <- with(VC.data, pairwise.t.test(length, dose, pool.sd = TRUE, 
                                              p.adjust.method = "none"))
OJ.crosstest$p.value
VC.crosstest$p.value
```

The only difference here is that we fail to reject the null hypothesis for the difference in tooth length between 1 and 2 mg dose for the Orange juice supplement.  

The second series of t-test, focusing on the difference in means between the two supplement types, is shown below
 in the hypothesis of equal variances. ($H_0$ : *"true difference in means is equal to 0"* 
versus $H_a$ : *"true difference in means is greater than 0"*). Tests are one sided with a default $\alpha = 5\%$ type 1 error rate.

```{r}

getp <- function(x) {
  a <- x[x$supplement == "OJ","length"]
  b <- x[x$supplement == "VC","length"]
  t.test(a, b, alternative = "greater", var.equal = TRUE)$p.value
  }

suptest <- ddply(ToothGrowth, .(dose), getp)
names(suptest) <- c("dose", "p-value")
round(suptest,4)
```

As we can see, assuming equal variances does not impact these results.

## pairwise t-tests with corrected p-values

Below are the results of pairwise t.test for each supplement type in the case where p-values are corrected by controlling the false discovery rate ("BH" correction).
The null hypothesis $H_0$ : *"true difference in means is equal to 0"* 
versus $H_a$ : *"true difference in means is not equal to 0"*,
Tests are two-sided with a default $\alpha = 5\%$ type 1 error rate and unequal variances.
```{r}
# Orange Juice supplement
OJ.data <- filter(ToothGrowth, supplement=="OJ")
OJ.crosstest <- with(OJ.data, pairwise.t.test(length, dose, pool.sd = FALSE, 
                                              p.adjust.method = "BH"))
# Ascorbic Acid supplement
VC.data <- filter(ToothGrowth, supplement=="VC")
VC.crosstest <- with(VC.data, pairwise.t.test(length, dose, pool.sd = FALSE, 
                                              p.adjust.method = "BH"))
OJ.crosstest$p.value
VC.crosstest$p.value
```

We see that the results are not changed compared to the unequal variances case and no correction (significant diffence in means is found in all cases).

